single_elimination

### 2025-05-01 mutual_information_curves.qmd
update 5/1/25

From Wikipedia, weight mutual information formula:

@@

@@

    # Loop through each row of prob_df
    for (j in 1:nrow(prob_df)) {
      # Calculate the squared difference between x_i and y_i for each row j
      diff <- abs(prob_df[[x_i]][j] - prob_df[[y_i]][j])
    w[j] <- w[j] + rank_weights[i]* diff^2  # Sum squared differences for each row
    }
  }
  w <- ifelse(w == 0.0000, 1, 1 / w)  # Inverse of squared differences and manually change 0 probabilities to 1 on the inverse
  w <- w/sum(w)
  # Compute weights dynamically for each row
  weighted_df <- prob_df %>%
    mutate(w = w)

  mutual_info <- factorial(num_teams) * sum(weighted_df$w * weighted_df$probability * log2(weighted_df$probability / (1/nrow(weighted_df) * weighted_df$probability)))

  # Return the weighted data frame with mutual information weights
  return(mutual_info)
}
mut_info(all_results_norm,8, c(10,5,1,0,0,0,0,0))
mut_info(de_norm_results_eight,8, c(10,5,1,0,0,0,0,0))
mut_info(test_rr_noTies,8, c(10,5,1,0,0,0,0,0))
mut_info(seven_games_se_results,8, c(10,5,1,0,0,0,0,0))

de_norm_results_eight <- de_norm_results %>% filter(teams == 8)
de_norm_results_eight$simulation <- rep(1:10000, each = nrow(de_norm_results_eight) / 10000)
all_results_norm$simulation <- rep(1:10000, each = nrow(all_results_norm) / 10000)
seven_games_se_results$simulation <- rep(1:10000, each = nrow(seven_games_se_results) / 10000)
# with randomly assigning ties
test_rr_noTies <- test_rr %>%
  group_by(simulation) %>%
  mutate(rank_hat = rank(rank_hat,ties.method="random"))

# Initialize variables
num_vars <- 8  # Total length of the vector
indices <- 1:num_vars  # Store the number of 1's in each step

# Function to compute mutual information for a given dataset
compute_mut_plots <- function(dataset) {
  results <- numeric(num_vars)
  for (i in 1:num_vars) {
    binary_vector <- c(rep(1, i), rep(0, num_vars - i))  # Construct vector
    results[i] <- mut_info(dataset, num_vars, binary_vector)  # Compute MI
  }
  return(results)
}

# Compute mutual information for each dataset
mi_de_norm <- compute_mut_plots(de_norm_results_eight)
mi_se_norm <- compute_mut_plots(all_results_norm)
mi_test_rr <- compute_mut_plots(test_rr_noTies)
mi_seven_se <- compute_mut_plots(seven_games_se_results)

# Plot results
plot(indices, mi_de_norm, type="b", pch=19, col="blue", ylim=range(c(mi_de_norm, mi_se_norm, mi_test_rr, mi_seven_se)),
     xlab="Number of 1's in the vector", ylab="Mutual Information Bits", main="Mutual Information vs. Number of 1's")

# Add other datasets
lines(indices, mi_se_norm, type="b", pch=19, col="red")
lines(indices, mi_test_rr, type="b", pch=19, col="green")
lines(indices, mi_seven_se, type = "b", pch=19, col="black")

# Add legend
legend("topleft", legend=c("Double Elimination", "Single Elimination", "Round Robin", "7 Game Single Elimination"), 
       col=c("blue", "red", "green", "black"), pch=19, lty=1)

### Single Elimination vs Double Elimination ###

# Plot results
plot(indices, mi_de_norm, type="b", pch=19, col="blue", ylim=range(c(mi_de_norm, mi_se_norm)),
     xlab="Number of 1's in the vector", ylab="Mutual Information Bits", main="Single Elimination vs Double Elimination")

# Add other datasets
lines(indices, mi_se_norm, type="b", pch=19, col="red")

# Add legend
legend("bottomright", legend=c("Double Elimination", "Single Elimination"), 
       col=c("blue", "red"), pch=19, lty=1)

```

>>>>

From Wikipedia, weighted mutual information formula:

@@

Rajski's distance: $1 - \frac{I(X,Y)}{H(X,Y)}$

Interpretation: the fraction of uncertainty not explained by mutual information

<https://en.wikipedia.org/wiki/Inverse_distance_weighting>

Every permutation is not possible to elimination tournaments, so should we divide by every possible permutation for 8 teams, or just every possible permutation in the tournament structure?

@@

    # Loop through each row of prob_df
    for (j in 1:nrow(prob_df)) {
      #x_vals <- as.numeric(prob_df[j, paste0("x", 1:num_teams)])
      #y_vals <- as.numeric(prob_df[j, paste0("y", 1:num_teams)])
      # Calculate the squared difference between x_i and y_i for each row j
      diff <- abs(prob_df[[x_i]][j] - prob_df[[y_i]][j])
      #rank_diff <- abs(rank(x_vals) - rank(y_vals))
      w[j] <- w[j] + rank_weights[i] * diff^2  # Sum squared differences for each row
      # Count inversions (pairs where order is flipped)
      #inversions <- sum(outer(x_vals, x_vals, ">") & outer(y_vals, y_vals, "<"))
      #w[j] <- sum(exp(rank_diff) - 1)
      #w[j] <- w[j] + rank_weights[i] * (exp(inversions) - 1)
      #w[j] <- cor(x_vals, y_vals, method = "spearman", use = "complete.obs")
      # Kendal's Tau as weights?
    }
  }
  #w <- (w + 1) / 2
  #w[w == Inf] <- 0
  w <- 1 / (w+1)  # Inverse of squared differences and manually change 0 probabilities to 1 on the inverse
  #w <- w/sum(w)
  # Compute weights dynamically for each row
  weighted_df <- prob_df %>%
    mutate(w = w)
  #print(weighted_df)

  numerator <- sum(weighted_df$w * weighted_df$probability * log2(weighted_df$probability / (1/factorial(num_teams) * (1/factorial(num_teams)))))
  denominator <- log2(factorial(num_teams))

  Raj_dist <- 1 - numerator/denominator

  # Return the weighted data frame with mutual information weights
  return(Raj_dist)
}

de_norm_results_eight <- de_norm_results %>% filter(teams == 8)
de_norm_results_eight$simulation <- rep(1:10000, each = nrow(de_norm_results_eight) / 10000)

all_results_norm$simulation <- rep(1:10000, each = nrow(all_results_norm) / 10000)
seven_games_se_results$simulation <- rep(1:10000, each = nrow(seven_games_se_results) / 10000)
# with randomly assigning ties
test_rr_noTies <- test_rr %>%
  group_by(simulation) %>%
  mutate(rank_hat = rank(rank_hat,ties.method="random"))
perms_list <- permn(8)
all_data <- list()
# Loop through each permutation and create the data frame
for (i in 1:length(perms_list)) {
# Create a temporary data frame for the current permutation
temp_df <- data.frame(
true_rank = rep(1:8, each = 1),  # Each row gets a true_rank from 1 to 8
rank_hat = perms_list[[i]],      # The current permutation's ranks
simulation = rep(i, 8)           # Assign the simulation number to each row
)
# Add the temporary data frame to the list
all_data[[i]] <- temp_df
}
# Combine all the data frames into one
perms_df <- do.call(rbind, all_data)

best <- data.frame(
  true_rank = rep(1:8, times = 10000),  # Repeat 1-8 for 10,000 simulations
  rank_hat = rep(1:8, times = 10000),   # Ensure rank_hat matches true_rank
  simulation = rep(1:10000, each = 8)   # Assign each set to a simulation
)
worst <- data.frame(
  true_rank = rep(1:8, times = 10000),  # Repeat 1-8 for 10,000 simulations
  rank_hat = rep(8:1, times = 10000),   # Ensure rank_hat matches true_rank
  simulation = rep(1:10000, each = 8)   # Assign each set to a simulation
)

# Function to compute mutual information for a given dataset
compute_mut_plots <- function(dataset) {
  num_teams <- length(unique(dataset$true_rank))
  results <- numeric(num_teams)
  for (i in 1:num_teams) {
    binary_vector <- c(rep(1, i), rep(0, num_teams - i))  # Construct vector
    results[i] <- mut_info(dataset, num_teams, binary_vector)  # Compute MI
  }
  return(results)
}

# Compute mutual information for each dataset
mi_de_norm <- compute_mut_plots(de_norm_results_eight)
mi_se_norm <- compute_mut_plots(all_results_norm)
mi_test_rr <- compute_mut_plots(test_rr_noTies)
mi_seven_se <- compute_mut_plots(seven_games_se_results)
mi_random <- compute_mut_plots(equal_strengths)
mi_se_unif <- compute_mut_plots(test_unif)
mi_bad <- compute_mut_plots(bad_seeding_structure)
mi_perms <- compute_mut_plots(perms_df)
mi_best <- compute_mut_plots(best)
mi_worst <- compute_mut_plots(worst)

mi_plots_df <- data.frame(
indices = 1:8,
"Double Elimination" = (1-mi_de_norm),
"Single Elimination" = (1-mi_se_norm),
"Equal Strengths Single Elimination" = (1-mi_random),
#"Unif_SE" = (1-mi_se_unif),
"Bad Seeding Structure Single Elimination" = (1-mi_bad),
"Every Possible Permutation" = (1-mi_perms),
"Round Robin" = (1-mi_test_rr),
"Worst Possible" = (1-mi_worst),
"Best Possible" = (1-mi_best)
)

# Reshape data into long format for ggplot
df_long <- pivot_longer(mi_plots_df, cols = -indices, names_to = "Method", values_to = "MI")
ggplot(df_long, aes(x = indices, y = MI, color = Method, group = Method)) +
geom_line() +           # Add lines
geom_point() +          # Add points
labs(
title = "Comparison of Tournament Structures with Normal Strengths",
x = "Top i Number of Teams",
y = "Mutual Information \\ Entropy (log10 scale)"
) +
theme_minimal() +       # Use a clean theme
theme(legend.position = "right") +
  scale_y_log10(
    labels = scales::label_number() 
  )

normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

# Apply normalization to all columns except "indices"
mi_norm_plot <- mi_plots_df
mi_norm_plot[ , -1] <- t(apply(mi_plots_df[ , -1], 1, normalize))
# Reshape data into long format for ggplot
df_long_norm <- pivot_longer(mi_norm_plot, cols = -indices, names_to = "Method", values_to = "MI")
windowsFonts(Arial = windowsFont("Arial"))
ggplot(df_long_norm, aes(x = indices, y = MI, color = Method, group = Method)) +
geom_line(linewidth = 2.5) +           # Add lines
geom_point(size=3) +          # Add points
labs(
title = "Comparison of Tournament Structures with Normal Strengths",
x = "Top i Number of Teams",
y = "Normalized Mutual Information \\ Entropy (log10 scale)"
) +
theme_minimal() +       # Use a clean theme
theme(text = element_text(family = "Arial", size = 32),
  legend.position = "bottom",
  legend.box = "vertical",
  plot.title = element_text(hjust = 0.5, size = 36),
  axis.title = element_text(size = 30),
  axis.text = element_text(size=30)) +
  guides(color = guide_legend(nrow = 4)) +
  scale_y_log10(
    labels = scales::label_number() 
  ) +
  scale_color_viridis_d(option = "H")
ggsave(
  filename = "tournament_comparison_plot.png",  # Change file type if needed
  plot = last_plot(),  # Saves the last ggplot created
  dpi = 600,        
  bg = "white",
  units = "in",
  width = 15.28,
  height = 13.43
)
```

include number of games (might be a range)

Single Elimination - third place and without third place; series = 3,5,7

Double Elimination

-   Repechage

-   True Double Elimination

-   Consolation Bracket (loser's bracket best place is third)

Triple Elimination

Deterministic Outcomes

Round Robin - with k rounds (1,2,3,5,7)

Stepladder - 8 vs 7, winner plays 6, etc.

Group Stage - winners of group advance, top 2 advance

Staged Round Robin/Jacobian Ladder - top 2 in each group form new group, bottom 2 new group (each new group is a new round robin)

High School Tennis Ladder - 1 vs 2, 3 vs 4, etc., Re-rank 2 vs 3, 4 vs 5, 6 vs 7, number of teams - 1 iterations

### 2025-05-01 single_elimination_shiny.qmd
update 5/1/25

  if (num_teams <= 3) {
    stop("Number of Teams must be greater than 3.")
  }
  teams <- paste(num_teams:1)
  if (log2(num_teams) %% 1 != 0) {
    power_of_2 <- 2^ceiling(log2(num_teams))
    teams <- c(teams, rep(NA, power_of_2 - num_teams))
  }
  if (distribution == "Normal"){
    strengths <- sapply(num_teams, function(n) { qnorm(1:n/(n+1)) })
    normal_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA,length(teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams))
    )
    df <- arrange(normal_strengths, true_rank)
  }
  else if (distribution == "Uniform"){
    strengths <- sapply(num_teams, function(n) { qunif(1:n/(n+1) , 0 , sqrt(12)) })
    unif_strength <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams)),
      rank_hat = rep(NA,length(teams))
    )
    df <- arrange(unif_strength, true_rank)
  }
  else if (distribution == "Manual") {
    cat("Please enter the strengths for each team in descending rank order:\n")
    strengths <- numeric(num_teams)
    for (i in num_teams:1) {
      repeat {
        input <- as.numeric(readline(paste("Strength for Team", num_teams - i + 1, ": ")))
        if (!is.na(input)) {
          strengths[i] <- input
          break
        } else {
          cat("Invalid input. Please enter a numeric value.\n")
        }
      }
    }
    manual_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA, length(teams)),
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams))
    )
    df <- arrange(manual_strengths, true_rank)
  }
  else {
    stop("Distribution not found: Enter Manual to input your own strengths")
  }

@@

test <- simulate_single_elimination_tournament(num_teams = 8, distribution = "Normal", ties=F, series=1, third_place=T)
test

@@

# Gets all results of a normal distribution num_teams between 4 and 64 single-game elimination
all_normal_results <- list()
for (teams in c(4,8,16,32,64)) {
  all_normal_results[[as.character(teams)]] <- list()
  for (replicate in 1:10000) {
    # Store each simulation result in the list with unique indexing
    all_normal_results[[as.character(teams)]][[replicate]] <- simulate_single_elimination_tournament(
      num_teams = teams,
      distribution = "Normal",
      ties = FALSE,
      series = 1,
      third_place = TRUE
    )
  }
}

@@

>>>>

  if (num_teams <= 3) {
    stop("Number of Teams must be greater than 3.")
  }
  teams <- paste(num_teams:1)
  if (log2(num_teams) %% 1 != 0) {
    power_of_2 <- 2^ceiling(log2(num_teams))
    teams <- c(teams, rep(NA, power_of_2 - num_teams))
  }
  if (distribution == "Normal"){
    strengths <- sapply(num_teams, function(n) { qnorm(1:n/(n+1)) })
    normal_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA,length(teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams))
    )
    df <- arrange(normal_strengths, true_rank)
  }
  else if (distribution == "Same"){
    strengths <- sapply(num_teams, function(n) {rep(0,n)})
    normal_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA,length(teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams))
    )
    df <- arrange(normal_strengths, true_rank)
  }
  else if (distribution == "Uniform"){
    strengths <- sapply(num_teams, function(n) { qunif(1:n/(n+1) , 0 , sqrt(12)) })
    unif_strength <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      game_wins = rep(0,length(teams)),
      game_losses = rep(0,length(teams)),
      rank_hat = rep(NA,length(teams))
    )
    df <- arrange(unif_strength, true_rank)
  }
  else if (distribution == "Manual") {
    cat("Please enter the strengths for each team in descending rank order:\n")
    strengths <- numeric(num_teams)
    for (i in num_teams:1) {
      repeat {
        input <- as.numeric(readline(paste("Strength for Team", num_teams - i + 1, ": ")))
        if (!is.na(input)) {
          strengths[i] <- input
          break
        } else {
          cat("Invalid input. Please enter a numeric value.\n")
        }
      }
    }
    manual_strengths <- data.frame(
      true_rank = as.numeric(teams),
      true_strength = c(strengths, rep(NA, length(teams) - num_teams)),
      rank_hat = rep(NA, length(teams)),
      game_wins = rep(0, length(teams)),
      game_losses = rep(0, length(teams))
    )
    df <- arrange(manual_strengths, true_rank)
  }
  else {
    stop("Distribution not found: Enter Manual to input your own strengths")
  }

@@

  test <- list()
for (teams in c(8)) {
  test[[as.character(teams)]] <- list()
  for (replicate in 1:10000) {
    # Store each simulation result in the list with unique indexing
    test[[as.character(teams)]][[replicate]] <- simulate_single_elimination_tournament(
      num_teams = teams,
      distribution = "Same",
      ties = FALSE,
      series = 1,
      third_place = TRUE
    )
  }
}
}
test <- simulate_single_elimination_tournament(num_teams = 8, distribution = "Normal", ties=F, series=1, third_place=T, seeding_structure = c(1,2,3,4,5,6,7,8))
test

@@

# Gets all results of a normal distribution num_teams between 4 and 64 single-game elimination
all_normal_results <- list()
for (teams in c(8)) {
  all_normal_results[[as.character(teams)]] <- list()
  for (replicate in 1:10000) {
    # Store each simulation result in the list with unique indexing
    all_normal_results[[as.character(teams)]][[replicate]] <- simulate_single_elimination_tournament(
      num_teams = teams,
      distribution = "Normal",
      ties = FALSE,
      series = 1,
      third_place = FALSE
    )
  }
}

@@

seven_games_se_results <- list()
for (teams in c(8)) {
  seven_games_se_results[[as.character(teams)]] <- list()
  for (replicate in 1:10000) {
    # Store each simulation result in the list with unique indexing
    seven_games_se_results[[as.character(teams)]][[replicate]] <- simulate_single_elimination_tournament(
      num_teams = teams,
      distribution = "Normal",
      ties = FALSE,
      series = 7,
      third_place = TRUE
    )
  }
}

seven_games_se_results <- do.call(rbind, lapply(names(seven_games_se_results), function(teams) {
  team_results <- do.call(rbind, seven_games_se_results[[teams]])
  team_results$teams <- as.numeric(teams)
  return(team_results)
}))
